Verified Nanopasses for Compiling Conditionals

### 2025-08-12 paper.tex
proof reading

The statement of the lemma for \lstinline{explicate-tail} is the most
straightforward. If expression \lstinline{e} compiles to
\lstinline{c}, and if \lstinline{e} evaluates to \lstinline{v}, then
\lstinline{c} also evaluates to \lstinline{v}:
\begin{lstlisting}
explicate-tail-correct :
  $∀${e}{$ρ$ $ρ'$}{s s'}{G G$′$}{c}{v}
  $→$ (s , $ρ$) $⊢$ e $⇓$ v $⊣$ (s' , $ρ'$)
  $→$ explicate-tail e G $≡$ (c , G$′$)
  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c $⇓$ v $⊣$ (s' , $ρ'$)
\end{lstlisting}
\noindent The cases in the proof of \lstinline{explicate-tail-correct}
for assignment and conditional expressions rely on the correctness
lemmas for \lstinline{explicate-assign} and
\lstinline{explicate-pred}.

The case for conditional expressions also two lemmas about the
monotonicity of the control flow graph.  The induction hypothesis for
\lstinline{e$₂$} says that the generated \lstinline{c$₂$} evaluates
using the CFG that was produced by the compilation of
\lstinline{e$₂$}, but to satisfy the induction hypothesis
for \lstinline{e$₁$} we need \lstinline{c$₂$} evaluated using the CFG
that was produced by the compilation of \lstinline{e$₃$}.
We define the following notation for
expression that one CFG is a prefix of another CFG:
\begin{lstlisting}
G$₁$ $↝$ G$₂$ = $∃$[ G ] G$₁$ ++ G $≡$ G$₂$
\end{lstlisting}
\noindent We prove that the output CFG of \lstinline{explicate-tail} is
an extension of its input CFG.
\begin{lstlisting}
explicate-tail-CFG : $∀${e}{G$₁$ G$₂$}{c}
  $→$ explicate-tail e G$₁$ $≡$ (c , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}
\noindent And we prove that the big-step semantics is monotonic with
respect to the $↝$ relation on control flow graphs.
\begin{lstlisting}
eval-CFG : $∀${c}{$ρ$ $ρ′$}{B$₁$ B$₂$}{s s$′$}{v}
  $→$ B$₁$ $↝$ B$₂$
  $→$ (s , $ρ$) , B$₁$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
  $→$ (s , $ρ$) , B$₂$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
\end{lstlisting}

@@

\begin{lstlisting}
explicate-pred-correct :
  $∀$ {e$₁$}{c$₁$ c$₂$ c}{b}{v}{s s$′$ s$″$}{$ρ$ $ρ′$ $ρ″$}{G G$′$}
  $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ Bool b $⊣$ (s$′$ , $ρ′$)
  $→$ ((s$′$,$ρ′$) , G $⊢ᶜ$ if b then c$₂$ else c$₃$ $⇓$ v $⊣$ (s$″$,$ρ″$))
  $→$ explicate-pred e$₁$ c$₂$ c$₃$ G $≡$ (c$₁$ , G$′$)
  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
\end{lstlisting}

\noindent This statement of correctness is interesting because it
requires the premise that either continuation \lstinline{c$₂$} or
\lstinline{c$₃$} evaluates to \lstinline{v}, depending on whether the
expression \lstinline{e$₁$} evaluates to \lstinline{true} or
\lstinline{false}. This explains the use of Agda's
\lstinline{if-then-else} in the second premise of
\lstinline{explicate-pred-correct}.\footnote{The formulation of this
lemma follows that of CompCert's \lstinline{transl\_condexpr\_prop}.}

The proof of \lstinline{explicate-pred-correct} uses many of the
monotonicity lemmas that we have already proved, plus one more.

@@

The responsibility of the Select Instructions pass is to convert the
assignment statements and primitive operations of the C-like language
into x86 instructions. The minor challenge is that many x86
instructions take just two arguments and have an update-in-place
semantics. The output of Select Instructions on the running example is
shown in figure~\ref{fig:running-example-x86}.\footnote{The redundant
\texttt{movq} instructions are removed by the Patch Instructions pass
in \emph{Essentials of Compilation}, which comes after Select
Instructions and Register Allocation.}

\begin{figure}[tp]
  \hrule
\begin{lstlisting}[mathescape=false]
    var x, y, t1, t2
    block_0:
       callq read_int
       movq %rax, x
       callq read_int
       movq %rax, y
       cmpq x, $0
       je block_1
       jmp block_3
    block_1:
        jmp block_2
    block_2:
        movq y, %rax
        subq $1, %rax
        movq %rax, %rax
    block_3:
        callq read_int
        movq %rax, t1
        cmpq t1, $1
        je block_4
        jmp block_5
    block_4:
        jmp block_2
    block_5:
        jmp block_6
    block_6:
        movq y, %rax
        subq $2, %rax
        movq %rax, t2
        movq t2, %rax
        subq x, %rax
        movq %rax, %rax
\end{lstlisting}
  \hrule
  \caption{Example after Select Instructions, in the x86 assembly language
    augmented with variables.}
\label{fig:running-example-x86}
\end{figure}

The function \lstinline{select-inst} translation a whole program in
the C-like language to x86.
\begin{lstlisting}
select-inst (Program n l B) =
  Program n l (map select-stmt B)
\end{lstlisting}

\noindent The \lstinline{select-stmt} function maps each statement to
one or more instructions. The return statement is compiled like an
assignment to the \lstinline{rax} register. (The calling convention
for x86 says that the return value of a function is passed in
\lstinline{rax}.)

@@

We compile the assignment of an equality comparison into a
\lstinline{CmpQ} and a \lstinline{MovQ}. The \lstinline{CmpQ} always
puts its result into \lstinline{rax}, so the \lstinline{MovQ} is
needed to place the result into the destination.\footnote{The real
\lstinline{cmpq} instruction is slightly stranger, involving
the \lstinline{EFLAGS} register.}

@@

When the statement is an assignment, the proof uses the following
lemma. It states that if expression \lstinline{e} evaluates to
\lstinline{v}, then the instructions generated for \lstinline{e} put
\lstinline{v} into the specified destination. The predicate
\lstinline{wrote} specifies the update to either the appropriate
variable or register depending on the kind of destination.

@@

\begin{lstlisting}
compile-correct : (e : Exp) (s : Inputs) (v : Value)
  $→$ interp-LIf e s $≡$ just v
  $→$ interp-x86-var (compile e) s v
compile-correct e s v ie =
   let i-rco = trans (rco-correct e s) ie in
   let i-lift = lift-locals-correct (rco e) s v i-rco in
   let i-exp = explicate-correct i-lift in
   select-inst-correct i-exp
\end{lstlisting}

@@

This paper takes a step in adapting the literature on compiler
correctness~\citep{Leroy:2006fe,Kumar:2014aa} to a university course
based on the incremental and nanopass approach of \emph{Essentials of
Compilation}~\citep{Siek:2023tr,Siek:2023ue}. We prove the correctness
of a compiler for a source language with let binding, integer
arithmetic, conditional expressions and Booleans, targeting an
x86-flavored register transfer language. The process of developing
mechanizing the proof in Agda encouraged a further split in the
nanopasses, separating Explicate Control into two passes, one that
lifts \lstinline{let} bindings and the second that compiles structured
control-flow into \lstinline{goto}. In designing these compiler passes
and proofs, it is enjoyable to trace the history of ideas going back
to \citet{Aho:1986qf}, \citet{Dybvig:1990aa}, and Olivier Danvy's
one-pass transformation into monadic normal form~\citep{Danvy:2003fk}.

>>>>

The statement of the lemma for \lstinline{explicate-tail} is the most
straightforward. If expression \lstinline{e} compiles to
\lstinline{c} and \lstinline{e} evaluates to \lstinline{v}, then
\lstinline{c} also evaluates to \lstinline{v}:
\begin{lstlisting}
explicate-tail-correct :
  $∀${e}{$ρ$ $ρ'$}{s s'}{G G$′$}{c}{v}
  $→$ (s , $ρ$) $⊢$ e $⇓$ v $⊣$ (s' , $ρ'$)
  $→$ explicate-tail e G $≡$ (c , G$′$)
  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c $⇓$ v $⊣$ (s' , $ρ'$)
\end{lstlisting}
\noindent The cases in the proof of \lstinline{explicate-tail-correct}
for assignment and conditional expressions rely on the correctness
lemmas for \lstinline{explicate-assign} and
\lstinline{explicate-pred}.

The case for conditional expressions requires two lemmas about the
monotonicity of the control flow graph.  The induction hypothesis for
\lstinline{e$₂$} says that the generated \lstinline{c$₂$} evaluates
using the CFG that was produced by the compilation of
\lstinline{e$₂$}, but to satisfy the induction hypothesis
for \lstinline{e$₁$} we need \lstinline{c$₂$} evaluated using the CFG
that was produced by the compilation of \lstinline{e$₃$}.
We define the following notation for
saying that one CFG is a prefix of another CFG:
\begin{lstlisting}
G$₁$ $↝$ G$₂$ = $∃$[ G ] G$₁$ ++ G $≡$ G$₂$
\end{lstlisting}
\noindent We prove that the output CFG of \lstinline{explicate-tail} is
an extension of its input CFG.
\begin{lstlisting}
explicate-tail-CFG : $∀${e}{G$₁$ G$₂$}{c}
  $→$ explicate-tail e G$₁$ $≡$ (c , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}
\noindent We prove that the big-step semantics is monotonic with
respect to the $↝$ relation on control flow graphs.
\begin{lstlisting}
eval-CFG : $∀${c}{$ρ$ $ρ′$}{B$₁$ B$₂$}{s s$′$}{v}
  $→$ B$₁$ $↝$ B$₂$
  $→$ (s , $ρ$) , B$₁$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
  $→$ (s , $ρ$) , B$₂$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
\end{lstlisting}

@@

\begin{lstlisting}
explicate-pred-correct :
  $∀$ {e$₁$}{c$₁$ c$₂$ c}{b}{v}{s s$′$ s$″$}{$ρ$ $ρ′$ $ρ″$}{G G$′$}
  $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ Bool b $⊣$ (s$′$ , $ρ′$)
  $→$ ((s$′$,$ρ′$) , G $⊢ᶜ$ (if b then c$₂$ else c$₃$) $⇓$ v $⊣$ (s$″$,$ρ″$))
  $→$ explicate-pred e$₁$ c$₂$ c$₃$ G $≡$ (c$₁$ , G$′$)
  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
\end{lstlisting}

\noindent This statement is interesting because it requires the
premise that either continuation \lstinline{c$₂$} or \lstinline{c$₃$}
evaluates to \lstinline{v}, depending on whether the expression
\lstinline{e$₁$} evaluates to \lstinline{true} or
\lstinline{false}. This explains the 
\lstinline{(if b then c$₂$ else c$₃$)} in the second premise.\footnote{The
formulation of this lemma follows that of CompCert's
\lstinline{transl\_condexpr\_prop}.}

The proof of \lstinline{explicate-pred-correct} uses many of the
monotonicity lemmas that we have already proved, plus the following
one.

@@

The responsibility of the Select Instructions pass is to convert the
assignment statements and primitive operations of the C-like language
into x86 instructions. The minor challenge is that many x86
instructions take only two arguments and have an update-in-place
semantics. The output of Select Instructions on the running example is
shown in figure~\ref{fig:running-example-x86}.\footnote{The redundant
\texttt{movq} instructions are removed by the Patch Instructions pass
in \emph{Essentials of Compilation}, which comes after Select
Instructions and Register Allocation.}

\begin{figure}[tp]
  \hrule
\begin{lstlisting}[mathescape=false,basicstyle=\ttfamily\footnotesize]
    var x, y, t1, t2
    block_0:
       callq read_int
       movq %rax, x
       callq read_int
       movq %rax, y
       cmpq x, $0
       je block_1
       jmp block_3
    block_1:
        jmp block_2
    block_2:
        movq y, %rax
        subq $1, %rax
        movq %rax, %rax
    block_3:
        callq read_int
        movq %rax, t1
        cmpq t1, $1
        je block_4
        jmp block_5
    block_4:
        jmp block_2
    block_5:
        jmp block_6
    block_6:
        movq y, %rax
        subq $2, %rax
        movq %rax, t2
        movq t2, %rax
        subq x, %rax
        movq %rax, %rax
\end{lstlisting}
  \hrule
  \caption{Example after Select Instructions, in the x86 assembly language
    augmented with variables.}
\label{fig:running-example-x86}
\end{figure}

The function \lstinline{select-inst} translates a whole program in
the C-like language to x86.
\begin{lstlisting}
select-inst (Program n l B) =
  Program n l (map select-stmt B)
\end{lstlisting}

\noindent The \lstinline{select-stmt} function maps each statement to
one or more instructions. The \lstinline{Return} statement is compiled
like an assignment to the \lstinline{rax} register. (The calling
convention for x86 says that the return value of a function is passed
in \lstinline{rax}.)

@@

We compile the assignment of an equality comparison into a
\lstinline{CmpQ} and a \lstinline{MovQ}. The \lstinline{CmpQ} puts its
result into \lstinline{rax}, so the \lstinline{MovQ} is needed to
place the result into the destination.\footnote{The real
\lstinline{cmpq} instruction is stranger still, involving the
\lstinline{EFLAGS} register.}

@@

When the statement is an assignment, the proof uses the following
lemma. It states that when expression \lstinline{e} evaluates to
\lstinline{v}, the instructions generated for \lstinline{e} put
\lstinline{v} into the specified destination. The predicate
\lstinline{wrote} specifies the update to either the appropriate
variable or register depending on the kind of destination.

@@

\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
compile-correct : (e : Exp) (s : Inputs) (v : Value)
  $→$ interp-LIf e s $≡$ just v
  $→$ interp-x86-var (compile e) s v
compile-correct e s v ie =
   let i-rco = trans (rco-correct e s) ie in
   let i-lift = lift-locals-correct (rco e) s v i-rco in
   let i-exp = explicate-correct i-lift in
   select-inst-correct i-exp
\end{lstlisting}

@@

This paper takes a step in adapting the literature on compiler
correctness~\citep{Leroy:2006fe,Kumar:2014aa} to a university course
based on the incremental and nanopass approach of \emph{Essentials of
Compilation}~\citep{Siek:2023tr,Siek:2023ue}. We prove the correctness
of a compiler for a source language with let binding, integer
arithmetic, conditional expressions and Booleans, targeting an
x86-flavored register transfer language. The process of developing
mechanizing the proof in Agda encouraged a further split in the
nanopasses, separating Explicate Control into two passes, one that
lifts \lstinline{let} bindings and the second that compiles structured
control-flow into \lstinline{goto}. In designing these compiler passes
and proofs, it is enjoyable to trace the history of ideas going back
to \citet{Aho:1986qf}, \citet{Dybvig:1990aa}, and Olivier Danvy's
transformation into monadic normal form~\citep{Danvy:2003fk}.
